Vegetation regrowth/encroachment (gjengroing)

### 2023-08-28 DRAFT_functional_plant_indicators_natopen.Rmd
analysis bit of natopen markdown

(2) GRUK monitoring data
GRUK stands for 'Grunnlendt åpen kalkmark', and represents a monitoring of open, limestone rich ecosystems with soils too shallow for forest establishment. The limestone rich focus of the scheme is motivated from a high biodiversity value and places all monitoring sites in an area around the Oslofjord. As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites.

In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

### Representativity in time and space
For semi-natural ecosystems, the ANO data in this analysis contain 233 plots in 83 sites, in principle distributed randomly across the country. As semi-natural ecosystems occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 233 plots are distributed across regions in the following way:

- Northern Norway: ?
- Central Norway:  ?
- Eastern Norway:  ?
- Western Norway:  ?
- Southern Norway: ?

For ASO, this analysis covers 381 meadow transects in 96 sites.

The 381 transects are distributed across regions in the following way:

- Northern Norway: ?
- Central Norway:  ?
- Eastern Norway:  ?
- Western Norway:  ?
- Southern Norway: ?

### Temporal coverage
The ANO evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme, as well as the 2022 ASO-data. Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al. 2018). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

@@

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.

@@

@@

ANO data

```{r, echo = F}
# Add data from cache
ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))

# st_layers(dsn = "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
# ANO.sp <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
#                    layer="ANO_Art")
# ANO.geo <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
#                   layer="ANO_SurveyPoint")
# head(ANO.sp)
# head(ANO.geo)
```

ASO data

```{r, echo = F}
# Add data from cache
ASO.sp<-readRDS(paste0(here::here(),"/data/cache/ASO.sp.RDS"))
ASO.geo<-readRDS(paste0(here::here(),"/data/cache/ASO.geo.RDS"))

#excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/ASO/Semi-naturlig_eng_S123_20#22.xlsx")

#ASO.sp <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/ASO/Semi-naturlig_eng_S123_2022.xlsx",sheet = "transektregistreringer_4")
#ASO.geo <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/ASO/Semi-naturlig_eng_S123_2022.xlsx",sheet = "surveyPoint_0")
#head(ASO.sp)
#head(ASO.geo)

Plant indicators from Tyler et al. (2021)
```{r, echo = F}
## Tyler indicator data
## data from cache
ind.dat <-readRDS(paste0(here::here(),"/data/cache/ind.dat.RDS"))

# ind.dat <- read.table("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
#                         sep = '\t', header=T, quote = '')
# head(ind.dat)

Generalized species lists NiN
```{r, echo = F}
# data from cache
Eco_State<-readRDS(paste0(here::here(),"/data/cache/Eco_State.RDS"))

# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
# str(Eco_State)
```

@@

### Plant indicator data
names(ind.dat)
names(ind.dat)[1] <- 'species'
ind.dat$species <- as.factor(ind.dat$species)
summary(ind.dat$species)
ind.dat <- ind.dat[!is.na(ind.dat$species),] # removing species-NAs
ind.dat[,'species.orig'] <- ind.dat[,'species'] # make a backup of the original species variable
ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2) # trimming away sub-species & co, and descriptor info

# the trimming above leaves some duplicates that need to be taken care of
ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
ind.dat <- ind.dat %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
                                                            "Anemone nemorosa x ranunculoides",
                                                            "Armeria maritima ssp. elongata",
                                                            "Asplenium trichomanes ssp. quadrivalens",
                                                            "Calystegia sepium ssp. spectabilis",
                                                            "Campanula glomerata 'Superba'",
                                                            "Dactylorhiza maculata ssp. fuchsii",
                                                            "Erigeron acris ssp. droebachensis",
                                                            "Erigeron acris ssp. politus",
                                                            "Erysimum cheiranthoides L. ssp. alatum",
                                                            "Euphrasia nemorosa x stricta var. brevipila",
                                                            "Galium mollugo x verum",
                                                            "Geum rivale x urbanum",
                                                            "Hylotelephium telephium (ssp. maximum)",
                                                            "Juncus alpinoarticulatus ssp. rariflorus",
                                                            "Lamiastrum galeobdolon ssp. argentatum",
                                                            "Lathyrus latifolius ssp. heterophyllus",
                                                            "Medicago sativa ssp. falcata",
                                                            "Medicago sativa ssp. x varia",
                                                            "Monotropa hypopitys ssp. hypophegea",
                                                            "Ononis spinosa ssp. hircina",
                                                            "Ononis spinosa ssp. procurrens",
                                                            "Pilosella aurantiaca ssp. decolorans",
                                                            "Pilosella aurantiaca ssp. dimorpha",
                                                            "Pilosella cymosa ssp. gotlandica",
                                                            "Pilosella cymosa ssp. praealta",
                                                            "Pilosella officinarum ssp. peleteranum",
                                                            "Poa x jemtlandica (Almq.) K. Richt.",
                                                            "Poa x herjedalica Harry Sm.",
                                                            "Ranunculus peltatus ssp. baudotii",
                                                            "Sagittaria natans x sagittifolia",
                                                            "Salix repens ssp. rosmarinifolia",
                                                            "Stellaria nemorum L. ssp. montana",
                                                            "Trichophorum cespitosum ssp. germanicum")
) )

# checking duplicates again
ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","Phosphorus","Grazing_mowing","Soil_disturbance","species.orig","species")]
# getting rid of sect. for Hieracium
ind.dat <- ind.dat %>% mutate(species=gsub("sect. ","",species.orig))
ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2)

ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","Phosphorus","Grazing_mowing","Soil_disturbance","species.orig","species")]
# only hybrids left -> get rid of these
ind.dat <- ind.dat[!duplicated(ind.dat[,'species']),]
ind.dat[duplicated(ind.dat[,'species']),"species"]

ind.dat$species <- as.factor(ind.dat$species)
summary(ind.dat$species)
# no duplicates left

head(ind.dat)

@@

## merge species data with indicators
ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                    y= ind.dat[,c("species", "Light","Moisture","Soil_reaction_pH","Nitrogen","Phosphorus","Grazing_mowing","Soil_disturbance")],
                    by.x="Species", by.y="species", all.x=T)
summary(ANO.sp.ind)
# checking which species didn't find a match
unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
# don't find synonyms for these in the ind lists

## trimming away the points without information on NiN, species or cover
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$Species),]
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$art_dekning),]

@@

### ASO data handling
## make ASO.geo into a spatial object
names(ASO.geo)
ASO.geo <- st_as_sf(x = ASO.geo, 
                        coords = c("x", "y"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84")

colnames(ASO.geo)
colnames(ASO.geo)[c(6,7,8,10,14,16,18,23,24)] <- c("Omradenummer_flatenummer","Eng_ID","ASO_ID","NiN_grunntype",
                                                   "bruksintensitet",
                                                   "beitetrykk",
                                                   "slatteintensitet",
                                                   "tungekjoretoy",
                                                   "slitasje")

## fix NiN-variables
# remove variable code in the data
ASO.geo$bruksintensitet
ASO.geo$bruksintensitet <- gsub("7JB-BA_", "", ASO.geo$bruksintensitet) 
unique(ASO.geo$bruksintensitet)
ASO.geo$bruksintensitet <- gsub("X", "NA", ASO.geo$bruksintensitet)
unique(ASO.geo$bruksintensitet)
ASO.geo$bruksintensitet <- as.numeric(ASO.geo$bruksintensitet)
unique(ASO.geo$bruksintensitet)

ASO.geo$beitetrykk
ASO.geo$beitetrykk <- gsub("7JB-BT_", "", ASO.geo$beitetrykk) 
unique(ASO.geo$beitetrykk)
ASO.geo$beitetrykk <- gsub("X", "NA", ASO.geo$beitetrykk)
unique(ASO.geo$beitetrykk)
ASO.geo$beitetrykk <- as.numeric(ASO.geo$beitetrykk)
unique(ASO.geo$beitetrykk)

ASO.geo$slatteintensitet
ASO.geo$slatteintensitet <- gsub("7JB-SI_", "", ASO.geo$slatteintensitet) 
unique(ASO.geo$slatteintensitet)
ASO.geo$slatteintensitet <- gsub("X", "NA", ASO.geo$slatteintensitet)
unique(ASO.geo$slatteintensitet)
ASO.geo$slatteintensitet <- as.numeric(ASO.geo$slatteintensitet)
unique(ASO.geo$slatteintensitet)

ASO.geo$tungekjoretoy
ASO.geo$tungekjoretoy <- gsub("MDirPRTK_", "", ASO.geo$tungekjoretoy) 
unique(ASO.geo$tungekjoretoy)
ASO.geo$tungekjoretoy <- gsub("X", "NA", ASO.geo$tungekjoretoy)
unique(ASO.geo$tungekjoretoy)
ASO.geo$tungekjoretoy <- as.numeric(ASO.geo$tungekjoretoy)
unique(ASO.geo$tungekjoretoy)

ASO.geo$slitasje
ASO.geo$slitasje <- gsub("MDirPRSE_", "", ASO.geo$slitasje) 
unique(ASO.geo$slitasje)
ASO.geo$slitasje <- gsub("X", "NA", ASO.geo$slitasje)
unique(ASO.geo$slitasje)
ASO.geo$slitasje <- as.numeric(ASO.geo$slitasje)
unique(ASO.geo$slitasje)

## fixing variable names and issues in ASO.sp
head(as.data.frame(ASO.sp))

colnames(ASO.sp)
colnames(ASO.sp)[8] <- "art_dekning"

# fix species names
ASO.sp <- as.data.frame(ASO.sp)

ASO.sp$Species <- ASO.sp$Navn
unique(as.factor(ASO.sp$Species))
ASO.sp$Species <- sub(".*?_", "", ASO.sp$Species) # lose the Norwegian name in the front
ASO.sp$Species <- gsub("_", " ", ASO.sp$Species) # replace underscore with space
ASO.sp$Species <- str_to_title(ASO.sp$Species) # make first letter capital
ASO.sp$Species <- gsub("( .*)","\\L\\1",ASO.sp$Species,perl=TRUE) # make capital letters after hyphon to lowercase
ASO.sp$Species <- gsub("( .*)","\\L\\1",ASO.sp$Species,perl=TRUE) # make capital letters after space to lowercase
ASO.sp[,'Species'] <- word(ASO.sp[,'Species'], 1,2) # lose subspecies

unique(as.factor(ASO.sp$Species))

## merge species data with indicators
ASO.sp.ind <- merge(x=ASO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                          y= ind.dat[,c("species","Light","Moisture","Soil_reaction_pH","Nitrogen","Phosphorus","Grazing_mowing","Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(ASO.sp.ind)

# checking which species didn't find a match
unique(ASO.sp.ind[is.na(ASO.sp.ind$Light & 
                          is.na(ASO.sp.ind$Nitrogen)),'Species'])

# fix species name issues

ASO.sp <- ASO.sp %>% 
  mutate(Species=str_replace(Species,"Cardamine dentata", "Cardamine pratensis")) %>%
  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecica")) %>%
  mutate(Species=str_replace(Species,"Chamerion angustifolium", "Chamaenerion angustifolium")) %>%
  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Vulgata")) %>%
  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
  mutate(Species=str_replace(Species,"Hylotelephium maximum", "Hylotelephium telephium")) %>%
  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
  mutate(Species=str_replace(Species,"Listera ovata", "Neottia ovata")) %>%
  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
  mutate(Species=str_replace(Species,"Potentilla anserina", "Argentina anserina")) %>%
  mutate(Species=str_replace(Species,"Rosa dumalis", "Rosa vosagiaca")) %>%
  mutate(Species=str_replace(Species,"Rubus fruticosus", "Rubus plicatus")) %>%
  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa"))

## merge species data with indicators
ASO.sp.ind <- merge(x=ASO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                    y= ind.dat[,c("species","Light","Moisture","Soil_reaction_pH","Nitrogen","Phosphorus","Grazing_mowing","Soil_disturbance")],
                    by.x="Species", by.y="species", all.x=T)
summary(ASO.sp.ind)
# checking which species didn't find a match
unique(ASO.sp.ind[is.na(ASO.sp.ind$Light & 
                          is.na(ASO.sp.ind$Nitrogen)),'Species'])

>>>>

(2) GRUK monitoring data
GRUK stands for 'Grunnlendt åpen kalkmark', and represents a monitoring of open, limestone rich ecosystems with soils too shallow for forest establishment. The limestone rich focus of the scheme is motivated from a high biodiversity value and places all monitoring sites in an area around the Oslofjord. As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites, all of which are mapped to the ecosystem type "åpen grunnlendt mark" (T2) and therein to the limestone-rich subtypes T2-C-7 and T2-C-8 in the Norwegian classification system for ecosystem types.
The vegetation analysis in GRUK record percent cover for every species of vascular plants in 1 sqm plots. In addition, a range of site variables related to ecosystem-specific drivers of state are recorded.
In the analysis in this document, we use all GRUK monitoring data data, which span the years 2020-22. The 2022-data also include an assessment of site condition, which is included in the analysis of results in this document. In the future, these data data may become available in Miljødirektoratets kartkatalog.

@@

(5) Grime's system of plant strategy scores (Grime 1974) comprises relative (too one another) scores for the competition-, stress-, and disturbance("ruderality")-related life strategy of plant species. In the analysis, in this document we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g. alien species for competition, climate change for stress, land use change for ruderality).

### Representativity in time and space
For nat-open ecosystems, the ANO data in this analysis contain 143 plots in 52 sites, in principle distributed randomly across the country. As nat-open ecosystems occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 143 plots are distributed across regions in the following way:

- Northern Norway: 29
- Central Norway:  39
- Eastern Norway:  34
- Western Norway:  24
- Southern Norway: 17

For GRUK, this analysis covers 1103 plots in 146 sites.

The 1103 plots are distributed across regions in the following way:

- Northern Norway: 0
- Central Norway:  0
- Eastern Norway:  1042
- Western Norway:  0
- Southern Norway: 61

### Temporal coverage
The ANO evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. GRUK covers 2020-2022. Thus, there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists for NiN ecosystem types (see also Töpper et al. 2018). For the nat-open ecosystem types these lists have been newly prepared by Evju et al. 2023. By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

@@

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.

@@

Grime J.P. 1974. Vegetation classification by reference to strategies. Nature 250(5461):26-31.

@@

ANO data: ANO.sp contains the species data, ANO.geo contains site data
```{r, echo = F}
# Add data from cache
ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))
```

GRUK data: GRUK.species contains the species data, GRUK.variable contains site data, GRUK2021.condition contains a field-based condition assessment from the 2021 season
```{r, echo = F}
# Add data from cache
GRUK.variables<-readRDS(paste0(here::here(),"/data/cache/GRUK.variables.RDS"))
GRUK.species<-readRDS(paste0(here::here(),"/data/cache/GRUK.species.RDS"))
GRUK2021.condition<-readRDS(paste0(here::here(),"/data/cache/GRUK2021.condition.RDS"))
```

Plant indicators from Tyler et al. (2021) and Grime (1974)
```{r, echo = F}
## data from cache
ind.Tyler<-readRDS(paste0(here::here(),"/data/cache/ind.Tyler.RDS"))
ind.Grime<-readRDS(paste0(here::here(),"/data/cache/ind.Grime.RDS"))
```

Generalized species lists (reference communities): natopen_NiN_ref contains the reference species lists, natopen_NiN_ref_spInfo contains additional taxonomic information for each species
```{r, echo = F}
# data from cache
natopen_NiN_ref<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref.RDS"))
natopen_NiN_ref_spInfo<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref_spInfo.RDS"))
```

@@

### Plant indicator data
# trimming away sub-species & co, and descriptor info
ind.Grime[,'species.orig'] <- ind.Grime[,'species']
ind.Grime[,'species'] <- word(ind.Grime[,'species'], 1,2)
names(ind.Tyler)[1] <- 'species'
ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
ind.Tyler <- ind.Tyler[!is.na(ind.Tyler$species),]
ind.Tyler[,'species.orig'] <- ind.Tyler[,'species']
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)

# dealing with 'duplicates'
ind.Grime[duplicated(ind.Grime[,'species']),"species"]
ind.Grime.dup <- ind.Grime[duplicated(ind.Grime[,'species']),c("species")]
ind.Grime[ind.Grime$species %in% ind.Grime.dup,]
# getting rid of the duplicates
ind.Grime <- ind.Grime %>% filter( !(species.orig %in% list("Carex viridula brachyrrhyncha",
                                                            "Dactylorhiza fuchsii praetermissa",
                                                            "Medicago sativa varia",
                                                            "Montia fontana chondrosperma",
                                                            "Papaver dubium lecoqii",
                                                            "Sanguisorba minor muricata")
) )
#ind.Grime[duplicated(ind.Grime[,'species']),"species"]

names(ind.Tyler)[1] <- 'species'
ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
ind.Tyler <- ind.Tyler[!is.na(ind.Tyler$species),]
ind.Tyler[,'species.orig'] <- ind.Tyler[,'species']
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)
#ind.Tyler2 <- ind.Tyler
#ind.Tyler <- ind.Tyler2
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
ind.Tyler <- ind.Tyler %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
                                                            "Anemone nemorosa x ranunculoides",
                                                            "Armeria maritima ssp. elongata",
                                                            "Asplenium trichomanes ssp. quadrivalens",
                                                            "Calystegia sepium ssp. spectabilis",
                                                            "Campanula glomerata 'Superba'",
                                                            "Dactylorhiza maculata ssp. fuchsii",
                                                            "Erigeron acris ssp. droebachensis",
                                                            "Erigeron acris ssp. politus",
                                                            "Erysimum cheiranthoides L. ssp. alatum",
                                                            "Euphrasia nemorosa x stricta var. brevipila",
                                                            "Galium mollugo x verum",
                                                            "Geum rivale x urbanum",
                                                            "Hylotelephium telephium (ssp. maximum)",
                                                            "Juncus alpinoarticulatus ssp. rariflorus",
                                                            "Lamiastrum galeobdolon ssp. argentatum",
                                                            "Lathyrus latifolius ssp. heterophyllus",
                                                            "Medicago sativa ssp. falcata",
                                                            "Medicago sativa ssp. x varia",
                                                            "Monotropa hypopitys ssp. hypophegea",
                                                            "Ononis spinosa ssp. hircina",
                                                            "Ononis spinosa ssp. procurrens",
                                                            "Pilosella aurantiaca ssp. decolorans",
                                                            "Pilosella aurantiaca ssp. dimorpha",
                                                            "Pilosella cymosa ssp. gotlandica",
                                                            "Pilosella cymosa ssp. praealta",
                                                            "Pilosella officinarum ssp. peleteranum",
                                                            "Poa x jemtlandica (Almq.) K. Richt.",
                                                            "Poa x herjedalica Harry Sm.",
                                                            "Ranunculus peltatus ssp. baudotii",
                                                            "Sagittaria natans x sagittifolia",
                                                            "Salix repens ssp. rosmarinifolia",
                                                            "Stellaria nemorum L. ssp. montana",
                                                            "Trichophorum cespitosum ssp. germanicum")
) )
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
# getting rid of sect. for Hieracium
ind.Tyler <- ind.Tyler %>% mutate(species=gsub("sect. ","",species.orig))
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)

#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
# only hybrids left -> get rid of these
ind.Tyler <- ind.Tyler[!duplicated(ind.Tyler[,'species']),]
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]

ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
# no duplicates left

# merge indicator data
ind.dat <- merge(ind.Grime,ind.Tyler, by="species", all=T)
#summary(ind.dat)
#ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat$species <- as.factor(ind.dat$species)
#summary(ind.dat$species)
head(ind.dat)

@@

## merge species data with indicators
ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                    y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                    by.x="Species", by.y="species", all.x=T)
summary(ANO.sp.ind)
# checking which species didn't find a match
unique(ANO.sp.ind[is.na(ANO.sp.ind$Light),'Species'])
# don't find synonyms for these in the ind lists

# trimming away the points without information on NiN, species or cover
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$Species),]
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$art_dekning),]

@@

### GRUK data handling
names(GRUK.variables)
names(GRUK.species)
names(GRUK2021.condition)

colnames(GRUK.species)[5] <- "art_dekning"

# fix species names
GRUK.species <- as.data.frame(GRUK.species)

GRUK.species$Species <- GRUK.species$Navn
unique(as.factor(GRUK.species$Species))
GRUK.species$Species <- sub(".*?_", "", GRUK.species$Species) # lose the Norwegian name in the front
GRUK.species$Species <- gsub("_", " ", GRUK.species$Species) # replace underscore with space
GRUK.species$Species <- str_to_title(GRUK.species$Species) # make first letter capital
GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after hyphon to lowercase
GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after space to lowercase
GRUK.species[,'Species'] <- word(GRUK.species[,'Species'], 1,2) # lose subspecies

## merge species data with indicators
GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID")], 
                          y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(GRUK.species.ind)

# checking which species didn't find a match
unique(GRUK.species.ind[is.na(GRUK.species.ind$Light & 
                                is.na(GRUK.species.ind$RR)),'Species'])

# fix species name issues
ind.dat <- ind.dat %>% 
  #  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
  #  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  #  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
  #  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
  #  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
  mutate(species=str_replace(species,"Rosa vosagica", "Rosa vosagiaca"))

GRUK.species <- GRUK.species %>% 
  mutate(Species=str_replace(Species,"Arabis wahlenbergii", "Arabis hirsuta")) %>%
  #  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
  #  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
  #  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
  #  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
  #  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
  #  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
  #  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  #  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
  #  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
  mutate(Species=str_replace(Species,"Cotoneaster scandinavicus", "Cotoneaster integerrimus")) %>%
  mutate(Species=str_replace(Species,"Cotoneaster symondsii", "Cotoneaster integrifolius")) %>%
  mutate(Species=str_replace(Species,"Cyanus montanus", "Centaurea montana")) %>%
  #  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
  mutate(Species=str_replace(Species,"Erysimum virgatum", "Erysimum strictum")) %>%
  #  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
  mutate(Species=str_replace(Species,"Festuca trachyphylla", "Festuca brevipila")) %>%
  #  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
  mutate(Species=str_replace(Species,"Hylotelephium maximum", "Sedum telephium")) %>%
  #  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
  #  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
  mutate(Species=str_replace(Species,"Lepidotheca suaveolens", "Matricaria discoidea")) %>%
  #  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
  mutate(Species=str_replace(Species,"Malus ×domestica", "Malus domestica")) %>%
  #  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
  #  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
  #  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
  #  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
